BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-30 arxiv/arxivSubmission_1/ms.tex
Updating

\begin{abstract}
We present \texttt{bhpwave}: a new Python-based, open-source tool for generating the gravitational waveforms of stellar-mass compact objects undergoing quasi-circular inspirals into rotating massive black holes. These binaries, known as extreme-mass-ratio inspirals (EMRIs), are exciting mHz gravitational wave sources for future space-based detectors such as LISA. Relativistic models of EMRI gravitational wave signals are necessary to unlock the full scientific potential of mHz detectors, yet few open-source EMRI waveform models exist. To help fill this void we built \texttt{bhpwave}, which uses the adiabatic approximation from black hole perturbation theory to rapidly construct gravitational waveforms based on the leading-order inspiral dynamics of the binary. In this work, we present the theoretical and numerical foundations underpinning \texttt{bhpwave}. We also demonstrate how \texttt{bhpwave} can be used to assess the impact of EMRI modeling errors on LISA gravitational wave data analysis. In particular, we find that for retrograde orbits and slowly-spinning black holes we can mismodel the gravitational wave phasing by as much as $\sim 10$ radians without significantly biasing EMRI parameter estimation.
\end{abstract}

@@

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (and possibly characterizing) EMRI signals \cite{GairJone06} and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing EMRIs \cite{HughETC21}.

@@

It is important to note that \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a variety of massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors} we use a few example problems to test the impact of modeling errors on parameter biases for EMRI data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. To support community collaboration and open-science, the codes for calculating all of the data, performing all of the analyses, and generating all of the plots presented in this work are made available via the public Github repository \texttt{bhpwave-article}.\footnote{\href{https://github.com/znasipak/bhpwave-article}{https://github.com/znasipak/bhpwave-article}} For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figures/domain_sampling.pdf}
    \caption{A $32\times 64$ grid in $(\alpha, \chi)$ mapped to the $(\hat{a}, \hat{\Omega})$ domain. Each dot represents a sampled point in the parameter space. These points are equally-spaced in $\alpha$ and in $\chi$, but cluster around $\hat{\Omega} = \hat{\Omega}_\mathrm{ISCO}$ and $\hat{a} = \hat{a}_\mathrm{max}$. The shading of each point reflects the relative density of points in the $(\hat{a}, \hat{\Omega})$ domain, while the dashed lines represent the domain boundaries.}
    \label{fig:domain}
\end{figure}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/orginal_parametrization.pdf}
    \caption{The energy flux $\mathcal{F}_E$ (left panel), (rescaled) orbital phase $\check{\Phi}$ (middle panel), and (rescaled) orbital time to merger $\check{t}$ as a function of (dimensionless) orbital frequency $\hat{\Omega}$. Different colors (shades) correspond with different values of $\hat{a}$. Therefore, lower spins terminate at lower ISCO frequencies. Note that the spin has a very small effect on the flux.}
    \label{fig:original}
\end{figure*}

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/rescaled_parametrization.pdf}
    \caption{The normalized energy flux $\mathcal{F}_N$ (left panel), normalized orbital phase $\check{\Phi}_N$ (middle panel), and normalized orbital time to merger $\check{t}_N$ as a function of normalized distance from the ISCO frequency $\hat{\Omega}_\mathrm{ISCO}$ in frequency space. We rescale each vertical axis by the leading post-Newtonian coefficient. Thus all curves asymptote to one at low frequencies.}
    \label{fig:rescaled}
\end{figure*}

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/final_parametrization.pdf}
    \caption{The normalized energy flux $\mathcal{F}_N$ (left panel), normalized orbital phase $\check{\Phi}_N$ (middle panel), and normalized orbital time to merger $\check{t}_N$ as a function of the final grid parameters $\alpha$ and $\chi$. The values of $\chi$ correspond to the same values of $\hat{a}$ used in Figure \ref{fig:original}.}
    \label{fig:final}
\end{figure*}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.99\linewidth]{figures/deltaPhiTest.pdf}
    \caption{The maximized phase difference $||\delta \Phi^I||_{\hat{a}, \hat{\Omega}}$ as described in Sec.~\ref{sec:traj} for different observation periods $T_\mathrm{obs}$. We can see that for any observation period, all the way up to 25 years, our phase spline maintains subradian phase accuracy when compared to direct numerical integration of the orbital phase.}
    \label{fig:obsTime}
\end{figure*}

@@

Similar to the trajectories, we presample the complex waveform amplitudes $B_{\ell m}$ on a fixed $65 \times 65$ grid in $(\alpha, \chi)$. To optimize the accuracy of our amplitude interpolation, we decompose $B_{\ell m}$ into its phase $\psi_{\ell m}$ and the log of its real amplitude $\ln A_{\ell m}$. We then interpolate these quantities independently. We generate mode data for $\ell \leq 20$ and $0 < m \leq \ell$. Even with the reduced sampling, the interpolation errors introduce estimated fractional errors $< 3\times 10^{-5}$ in $A_{\ell m}$ and absolute errors $<2 \times 10^{-5}$ for $\psi_{\ell m}$, as discussed in Appendix \ref{app:tests}.

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency, but for $\hat{a} \gtrsim 0.995$ the amplitudes can peak well before reaching the ISCO. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO, which is highlighted in the middle panel of \ref{fig:amplitudeVariation} for all $m =2$ modes. This behavior is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. Furthermore, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 15$ modes for a given orbital frequency. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot the $A_{\ell, m = \ell}$ amplitudes as functions of $r_0$ for different values of $\hat{a}$. We see that $|A_{15,15}/A_{2,2}|^2 \lesssim 5\times 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 15$ mode at a fixed frequency. For most year-long EMRI signals, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 15$ mode is $\lesssim 10^{-6}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell = m = 20$, even for rapidly-spinning black holes.

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/amp_combined.pdf}
    \caption{The top panel plots the waveform amplitude $A_{\ell=2, m=2}$ as a function of the radius of the orbit $r_0/M$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The bottom panel plots the $A_{\ell, m=\ell}$ modes as a function of $r_0/M$ for $\hat{a} = 0.9999$ and all $2 \leq \ell \leq 15$. The magnitude of the $\ell = m$ modes is ordered inversely with $\ell$ in the bottom panel. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 15$ being the mode with the smallest magnitude. The bottom (yellow) line in the top panel is identical to the top (yellow) line in the bottom panel.}
    \label{fig:amplitudeVariation}
\end{figure}

@@

Given a fixed time step $dt$, signal duration $T$, and binary source parameters (e.g., $M$, $\mu$, $r_0$, $d_L$), time-domain waveforms are constructed using \eqref{eqn:adiabaticWaveform}, \eqref{eqn:SSBtoSource}, \eqref{eqn:polarization}, with all functions evaluated from the trajectory and mode amplitude splines outlined above. We simplify the sum over $(\ell, m)$-modes using the identities $B_{\ell -m} = (-1)^\ell B^*_{\ell m}$ and
\begin{subequations}
    \begin{align}
    Y_{s\ell -m}(\theta) &= (-1)^{m} Y_{-s \ell m}(\theta),
    \\
    &= (-1)^{\ell} Y_{s \ell m}(\pi-\theta),
    \end{align}
\end{subequations}
leading to the reduced sums
\begin{subequations} \label{eqn:timeDomainReduced}
    \begin{align}
    h_+(t) = +\sum_{\ell=2}^\infty \sum_{ m = 1}^{\ell} \mathcal{A}_{\ell m} Y^+_{\ell m} \cos(\psi_{\ell m} + m[\phi - \Phi]),
    \\
    h_\times(t) = -\sum_{\ell=2}^\infty \sum_{ m = 1}^{\ell}  \mathcal{A}_{\ell m} Y^\times_{\ell m} \sin(\psi_{\ell m} + m[\phi - \Phi]),
\end{align}
\end{subequations}
where $\mathcal{A} = \mu A_{\ell m}/r$ and
\begin{subequations}
    \begin{align}
    Y^{+/\times}_{\ell m}(\theta) &= Y_{-2\ell m}(\theta) \pm Y_{-2\ell m}(\pi - \theta),
    \\
    &= Y_{-2\ell m}(\theta) \pm (-1)^{\ell + m} Y_{+2\ell m}(\theta).
    \end{align}
\end{subequations}

The truncation of the $(\ell, m)$-mode sum is determined by the power in each harmonic mode,
\begin{subequations} \label{eqn:powerLM}
    \begin{align}
        P_{\ell m} &= \int_0^T \left[ \left|\mathcal{A}_{\ell m}(t) Y_{\ell m}^+\right|^2 + \left|\mathcal{A}_{\ell m}(t) Y_{\ell m}^\times\right|^2  \right]dt ,
        \\
        &\approx \left[\left|Y_{\ell m}^+\right|^2 + \left|Y_{\ell m}^\times\right|^2 \right] \sum_{n=1}^{N} |\mathcal{A}_{\ell m}(\hat{\Omega}_n)|^2 \Delta t_n,
    \end{align}
\end{subequations}
where $\hat{\Omega}_n = \hat{\Omega}(t=0) + n \Delta\hat{\Omega}$, with orbital frequency spacing $\Delta \hat{\Omega} = [\hat{\Omega}(t=T) - \hat{\Omega}(t=0)]/N$, time spacing $\Delta t_n = t(\hat{\Omega}_n) - t(\hat{\Omega}_{n-1})$, and signal duration $T$. In this work, we find that $N = 500$ sufficiently approximates the mode power. Additionally, we find that an equal spacing in frequency space is more numerically efficient than an equal sampling in time. We include all $(\ell, m)$-modes that satisfy the selection criteria
\begin{align} \label{eqn:modeSelectionCriteria}
    {P_{\ell m}} > \epsilon_\mathrm{mode} \times {\sum_{\ell' = 2}^{\ell} \sum_{m'=1}^{\ell'}P_{\ell' m'}},
\end{align}
for a user-specified threshold $\epsilon_\mathrm{mode}$. Rather than computing the power for all of the modes and then removing the modes that do not satisfy \eqref{eqn:modeSelectionCriteria}, we instead perform a serial search. First we find all modes that meet \eqref{eqn:modeSelectionCriteria} for $m \leq 2$. We then increase $m$ to $m = 3$ and increment over $\ell$, beginning with $\ell = m$ until \eqref{eqn:modeSelectionCriteria} is no longer satisfied. We repeat this process of increasing $m$ and then $\ell$, until either $P_{mm}$ no longer satisfies \eqref{eqn:modeSelectionCriteria} or $m > 20$.

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/few_bhpwave_overlay.pdf}
    \includegraphics[width=0.98\linewidth]{figures/few_phase_comparison.pdf}
    \caption{Comparisons between waveforms produced by the FEW and \texttt{bhpwave} models for an EMRI with intrinsic parameters $(M, \mu, a, p_0, e_0, x) = (10^6 M_\odot, 10 M_\odot, 0, 12.05, 0, 1)$. The top panels show the evolution of the strain $h_+$ with time $t$. In the top plots, the solid (blue) lines correspond to the waveform computed by FEW, while the dashed (orange) lines are produced by \texttt{bhpwave}. The three different panels focus on three small time windows within the full four year signal. The bottom panel plots the difference in the gravitational wave phases between the waveform models. The solid (cyan) line shows the phase difference when the waveforms are evaluated at the same set of parameters. The dashed (black) line is the phase difference after rescaling the FEW parameters to account for different definitions of $GM_\odot$.}
    \label{fig:compareFEW}
\end{figure*}

@@

\begin{figure}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/dft_comparison.pdf}
    \caption{Comparison between the discrete Fourier transform of the time-domain gravitational wave strain $h_+$ produced by \texttt{bhpwave} (solid blue line) and the frequency-domain waveform for $h_+$ produced by \texttt{bhpwave} using the stationary phase approximation (dashed orange line). For reference, the sensitivity curve $S_n$ is also plotted as the dot-dashed (black) curve. Note that we downsample the number of the frequency samples included in this plot by a factor of 10000 in order to more clearly see how well the two waveforms agree, even in the highly oscillatory region between $10^{-3} - 10^{-2}$ Hz.}
    \label{fig:dft}
\end{figure}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/mcmc.pdf}
    \caption{Posterior distribution based on an parallel-tempered MCMC sampling (dark blue lines) and a Fisher matrix analysis (lighter orange lines).}
    \label{fig:mcmcOverlay}
\end{figure*}

>>>>

\begin{abstract}
We present \texttt{bhpwave}: a new Python-based, open-source tool for generating the gravitational waveforms of stellar-mass compact objects undergoing quasi-circular inspirals into rotating massive black holes. These binaries, known as extreme-mass-ratio inspirals (EMRIs), are exciting mHz gravitational wave sources for future space-based detectors such as LISA. Relativistic models of EMRI gravitational wave signals are necessary to unlock the full scientific potential of mHz detectors, yet few open-source EMRI waveform models exist. Thus we built \texttt{bhpwave}, which uses the adiabatic approximation from black hole perturbation theory to rapidly construct gravitational waveforms based on the leading-order inspiral dynamics of the binary. In this work, we present the theoretical and numerical foundations underpinning \texttt{bhpwave}. We also demonstrate how \texttt{bhpwave} can be used to assess the impact of EMRI modeling errors on LISA gravitational wave data analysis. In particular, we find that for retrograde orbits and slowly-spinning black holes we can mismodel the gravitational wave phasing by as much as $\sim 10$ radians without significantly biasing EMRI parameter estimation.
\end{abstract}

@@

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (and possibly characterizing) EMRI signals \cite{GairJone06} and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing\footnote{In a frame where the massive black hole's angular momentum is held fixed, precession is instead described in terms of the inclination of the small body relative to the equatorial plane of the more massive black hole.} EMRIs \cite{HughETC21}.

@@

It is important to note that \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a variety of massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin (which is a post-adiabatic effect) \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors} we present example problems for testing the impact of modeling errors on parameter biases for LISA data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. To support community collaboration and open-science, the codes for calculating all of the data, performing all of the analyses, and generating all of the plots presented in this work are made available via the public Github repository \texttt{bhpwave-article}.\footnote{\href{https://github.com/znasipak/bhpwave-article}{https://github.com/znasipak/bhpwave-article}} For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{domain_sampling.pdf}
    \caption{A $32\times 64$ grid in $(\alpha, \chi)$ mapped to the $(\hat{a}, \hat{\Omega})$ domain.  Each dot represents a sampled point in the parameter space. These points are equally-spaced in $\alpha$ and in $\chi$, but cluster around $\hat{\Omega} = \hat{\Omega}_\mathrm{ISCO}$ and $\hat{a} = \hat{a}_\mathrm{max}$. The shading of each point reflects the relative density of points in the $(\hat{a}, \hat{\Omega})$ domain, while the dashed lines represent the domain boundaries. Note that this is downsampled version of the numerical grid used for interpolating flux and trajectory information within \texttt{bhpwave}.}
    \label{fig:domain}
\end{figure}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.95\linewidth]{orginal_parametrization.pdf}
    \caption{The energy flux $\mathcal{F}_E$ (left panel), (rescaled) orbital phase $\check{\Phi}$ (middle panel), and (rescaled) orbital time to merger $\check{t}$ as a function of (dimensionless) orbital frequency $\hat{\Omega}$. Different colors (shades) correspond with different values of $\hat{a}$. Therefore, lower spins terminate at lower ISCO frequencies. Note that the spin has a very small effect on the flux.}
    \label{fig:original}
\end{figure*}

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{rescaled_parametrization.pdf}
    \caption{The normalized energy flux $\mathcal{F}_N$ (left panel), normalized orbital phase $\check{\Phi}_N$ (middle panel), and normalized orbital time to merger $\check{t}_N$ as a function of normalized distance from the ISCO frequency $\hat{\Omega}_\mathrm{ISCO}$ in frequency space. We rescale each vertical axis by the leading post-Newtonian coefficient. Thus all curves asymptote to one at low frequencies.}
    \label{fig:rescaled}
\end{figure*}

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{final_parametrization.pdf}
    \caption{The normalized energy flux $\mathcal{F}_N$ (left panel), normalized orbital phase $\check{\Phi}_N$ (middle panel), and normalized orbital time to merger $\check{t}_N$ as a function of the final grid parameters $\alpha$ and $\chi$. The values of $\chi$ correspond to the same values of $\hat{a}$ used in Figure \ref{fig:original}.}
    \label{fig:final}
\end{figure*}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.99\linewidth]{deltaPhiTest.pdf}
    \caption{The maximized phase difference $||\delta \Phi^I||_{\hat{a}, \hat{\Omega}}$ as described in Sec.~\ref{sec:traj} for different observation periods $T_\mathrm{obs}$. We can see that for any observation period, all the way up to 25 years, our phase spline maintains subradian phase accuracy when compared to direct numerical integration of the orbital phase.}
    \label{fig:obsTime}
\end{figure*}

@@

Similar to the trajectories, we presample the complex waveform amplitudes $B_{\ell m}$ on a fixed $65 \times 65$ grid in $(\alpha, \chi)$. To optimize the accuracy of our amplitude interpolation, we decompose $B_{\ell m}$ into its phase $\psi_{\ell m}$ and the log of its real amplitude $\ln A_{\ell m}$. We then interpolate these quantities independently. We generate mode data for $\ell \leq 15$ and $0 < m \leq \ell$. Even with the reduced sampling, the interpolation errors introduce estimated fractional errors $< 3\times 10^{-5}$ in $A_{\ell m}$ and absolute errors $<2 \times 10^{-5}$ for $\psi_{\ell m}$, as discussed in Appendix \ref{app:tests}.

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency. For $\hat{a} \gtrsim 0.995$ the amplitudes can instead peak well before reaching the ISCO and then decay as the orbital frequency increases. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO, which is highlighted in the middle panel of Figure \ref{fig:amplitudeVariation} for all $m =2$ modes. This behavior is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. As a result, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 15$ modes for a given orbital frequency. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot the $A_{\ell, m = \ell}$ amplitudes as functions of $r_0$ for different values of $\hat{a}$. We see that $|A_{15,15}/A_{2,2}|^2 \lesssim 5\times 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 15$ mode at a fixed frequency. For most year-long EMRI signals, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 15$ mode is $\lesssim 10^{-6}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell \leq 15$.

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{amp_combined.pdf}
    \caption{Waveform amplitudes $A_{\ell, m}$ as a function of the radius of the orbit $r_0/M$. The top panel plots $A_{\ell=2, m=2}$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The middle and bottom panels plot $A_{\ell, m=2}$ and $A_{\ell, m=\ell}$, respectively, for $\hat{a} = 0.9999$ and $2 \leq \ell \leq 15$. In the two lower panels, the mode magnitudes are ordered inversely with $\ell$. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 15$ being the mode with the smallest magnitude (and darkest solid line). }
    %The bottom (yellow) line in the top panel is identical to the top (yellow) lines in the lower two panels.
    \label{fig:amplitudeVariation}
\end{figure}

@@

Given a fixed time step $dt$, signal duration $T$, and binary source parameters (e.g., $M$, $\mu$, $r_0$, $d_L$), time-domain waveforms are constructed using \eqref{eqn:adiabaticWaveform}, \eqref{eqn:SSBtoSource}, \eqref{eqn:polarization}, with all functions evaluated from the trajectory and mode amplitude splines outlined above. We simplify the sum over $(\ell, m)$-modes using $B_{\ell -m} = (-1)^\ell B^*_{\ell m}$ and the identities of spin-weighted spherical harmonics,
\begin{subequations}
    \begin{align}
    Y_{s\ell -m}(\theta) &= (-1)^{m} Y_{-s \ell m}(\theta),
    \\
    &= (-1)^{\ell} Y_{s \ell m}(\pi-\theta),
    \end{align}
\end{subequations}
leading to the reduced sums
\begin{subequations} \label{eqn:timeDomainReduced}
    \begin{align}
    h_+(t) = +\sum_{\ell=2}^\infty \sum_{ m = 1}^{\ell} \mathcal{A}_{\ell m} Y^+_{\ell m} \cos(\psi_{\ell m} + m[\phi - \Phi]),
    \\
    h_\times(t) = -\sum_{\ell=2}^\infty \sum_{ m = 1}^{\ell}  \mathcal{A}_{\ell m} Y^\times_{\ell m} \sin(\psi_{\ell m} + m[\phi - \Phi]),
\end{align}
\end{subequations}
where $\mathcal{A}_{\ell m} = \mu A_{\ell m}/r$ and
\begin{subequations}
    \begin{align}
    Y^{+/\times}_{\ell m}(\theta) &= Y_{-2\ell m}(\theta) \pm Y_{-2\ell m}(\pi - \theta),
    \\
    &= Y_{-2\ell m}(\theta) \pm (-1)^{\ell + m} Y_{+2\ell m}(\theta).
    \end{align}
\end{subequations}

The truncation of the $(\ell, m)$-mode sum is determined by the power in each harmonic mode,
\begin{subequations} \label{eqn:powerLM}
    \begin{align}
        P_{\ell m} &= \int_0^T \left[ \left|\mathcal{A}_{\ell m}(t) Y_{\ell m}^+\right|^2 + \left|\mathcal{A}_{\ell m}(t) Y_{\ell m}^\times\right|^2  \right]dt ,
        \\
        &\approx \left[\left|Y_{\ell m}^+\right|^2 + \left|Y_{\ell m}^\times\right|^2 \right] \sum_{n=1}^{N} |\mathcal{A}_{\ell m}(\hat{\Omega}_n)|^2 \Delta t_n,
    \end{align}
\end{subequations}
where $\hat{\Omega}_n = \hat{\Omega}(t=0) + n \Delta\hat{\Omega}$, with orbital frequency spacing $\Delta \hat{\Omega} = [\hat{\Omega}(t=T) - \hat{\Omega}(t=0)]/N$, time spacing $\Delta t_n = t(\hat{\Omega}_n) - t(\hat{\Omega}_{n-1})$, and signal duration $T$. In this work, we find that $N = 500$ sufficiently approximates the mode power. Additionally, we find that an equal spacing in frequency space is more numerically efficient than an equal sampling in time. We include all $(\ell, m)$-modes that satisfy the selection criteria
\begin{align} \label{eqn:modeSelectionCriteria}
    {P_{\ell m}} > \epsilon_\mathrm{mode} \times {\sum_{\ell' = 2}^{\ell} \sum_{m'=1}^{\ell'}P_{\ell' m'}},
\end{align}
for a user-specified threshold $\epsilon_\mathrm{mode}$. Rather than computing the power for all of the modes and then removing the modes that do not satisfy \eqref{eqn:modeSelectionCriteria}, we instead perform a serial search. First we find all modes that meet \eqref{eqn:modeSelectionCriteria} for $m \leq 2$. We then increase $m$ to $m = 3$ and increment over $\ell$, beginning with $\ell = m$ until \eqref{eqn:modeSelectionCriteria} is no longer satisfied. We repeat this process of increasing $m$ and then $\ell$, until either $P_{mm}$ no longer satisfies \eqref{eqn:modeSelectionCriteria} or $m > \ell_\mathrm{max} = 15$.

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{few_bhpwave_overlay.pdf}
    \includegraphics[width=0.98\linewidth]{few_phase_comparison.pdf}
    \caption{Comparisons between waveforms produced by the FEW and \texttt{bhpwave} models for an EMRI with intrinsic parameters $(M, \mu, a, p_0, e_0, x) = (10^6 M_\odot, 10 M_\odot, 0, 12.05, 0, 1)$. The top panels show the evolution of the strain $h_+$ with time $t$. In the top plots, the solid (blue) lines correspond to the waveform computed by FEW, while the dashed (orange) lines are produced by \texttt{bhpwave}. The three different panels focus on three small time windows within the full four year signal. The bottom panel plots the difference in the gravitational wave phases between the waveform models. The solid (cyan) line shows the phase difference when the waveforms are evaluated at the same set of parameters. The dashed (black) line is the phase difference after rescaling the FEW parameters to account for different definitions of $GM_\odot$.}
    \label{fig:compareFEW}
\end{figure*}

@@

\begin{figure}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{dft_comparison.pdf}
    \caption{Comparison between the discrete Fourier transform of the time-domain gravitational wave strain $h_+$ produced by \texttt{bhpwave} (solid blue line) and the frequency-domain waveform for $h_+$ produced by \texttt{bhpwave} using the stationary phase approximation (dashed orange line). For reference, the sensitivity curve $S_n$ is also plotted as the dot-dashed (black) curve. Note that we downsample the number of the frequency samples included in this plot by a factor of 10000 in order to more clearly see how well the two waveforms agree, even in the highly oscillatory region between $10^{-3} - 10^{-2}$ Hz.}
    \label{fig:dft}
\end{figure}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{mcmc.pdf}
    \caption{Posterior distribution based on an parallel-tempered MCMC sampling (dark blue lines) and a Fisher matrix analysis (lighter orange lines).}
    \label{fig:mcmcOverlay}
\end{figure*}
